---
title: "generate_spacebreak examples"
author: "Clay Morrow"
date: "7/14/2022"
output: 
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(breakR)
library(dplyr)
library(ggplot2)
```

# Component functions

## Piecewise time series

In order to facilitate time-dependent trends, `breakR` uses piecewise time 
series functions to generate temporal data. The way these functions are created
are with breakpoint data frames, generated with `break_frame()`. For example,
if we want a time series from 0 to 10, with an increasing trend from 0-5 and 
a decreasing trend from 5-9, we would use the following code and resulting 
breakpoint data frame:

```{r}
(break.df <- break_frame(nodes = c(0, 5, 9), slopes = c(.2, -.2)))
```

To build the piecewise function from this output, the `breakwise()` function
is used. Here, we'll evaluate the piecewise function for each of the ten time 
points (0-9).

```{r}
(f.t <- breakwise(0:9, break.df))
```

And here's what that looks like as a time series plot:

```{r}
plot(x = 0:9, y = f.t, xlab = 'time')
```

With these functions, we can make simple trends:

```{r}
df = break_frame(nodes = c(0, 49), slopes = .1)
plot(x = 0:49, y = breakwise(0:49, df), xlab = "time")
```

or very complicated time series with multiple breakpoints:

```{r}
df = break_frame(nodes = c(0, 10, 11, 20, 35, 49), slopes = c(.1, -1, 0, .05, -.01))
plot(x = 0:49, y = breakwise(0:49, df), type = "line", xlab = "time")
```

## Seasonality

We can also add seasonality with the `season` function. For example, if we want 
a 3-year monthly time series, with a 12-month seasonal cycle, we would use the following

```{r}
seas = season(t = seq(0, 2, by = 1/12), amp = 1, per = 1)
plot(y = seas, x = seq(0, 2, by = 1/12), type = "b")
```

And we can even add random variation to this pattern:

```{r}
seas = season(t = seq(0, 2, by = 1/12), amp = 1, per = 1, noise.sd = .2)
plot(y = seas, x = seq(0, 2, by = 1/12), type = "b")
```

## Piecewise functions with seasonality

Making a time trend with seasonality is as simple as adding these functionalities
together.

```{r}
time = seq(0, 9, by = 1/12)
breaks = c(0, 3, 5, 9)
df = break_frame(nodes = breaks, slopes = c(.1, 0, -.1))
f.t = breakwise(time, df)
seas = season(t = time, amp = .05, per = 1)
y = f.t + seas

plot(y = y, x = time)
lines(x = time, y = f.t, lty = 2, col = "blue")
abline(v = breaks, lty = 2, col = "grey70")
```

## Spatial variation

The function `generate_spcov()` is used to create spatial relationships. For 
example, if we want a 25 pixel map with spatial autocorrelation that ranges
across half of the map, we can use 

```{r}
space = generate_spcov(map.dims = c(3, 3), covar.pars = list(range = 1))
```

This function creates spatial coordinates for all the pixels:

```{r}
space$coord.df
```

a distance matrix:

```{r}
round(space$D.mat, 1)
```

and a covariances matrix: 

```{r}
round(space$covar.mat, 1)
```

A random spatial variable can be created with `rnorm_spcor()`. We can use the 
covariance structure created above to create 5 random samples for each pixel. 

```{r}
rnorm_spcor(n = 5, covar = space$covar.mat)
```

## Spatiotemporal data

And finally, we can generate a spatiotemporal variable with the `sparima_sim()`
function. Here, we create a spatiotemporal variable with 20% AR(1) temporal 
autocorrelation, and the spatial covariance generated earlier, over 6 time
points:

```{r}
(spacetime = sparima_sim(ntime = 6, covar = space$covar.mat, model = list(ar = .2)))
```

This is what those data look like visually:

```{r}
data.frame(space$coord.df, Y = spacetime) %>% 
  reshape2::melt(id.vars = c("pixel", "coords.x", "coords.y")) %>% 
  ggplot(aes(x = coords.x, y = coords.y, fill = value)) + 
  geom_tile() + 
  facet_wrap(~variable)
```

And here they are when we add an underlying trend to the data:

```{r}
st.trend = t(apply(spacetime, MARGIN = 1, FUN = function(x){x + .8*0:5}))

data.frame(space$coord.df, Y = st.trend) %>% 
  reshape2::melt(id.vars = c("pixel", "coords.x", "coords.y")) %>% 
  ggplot(aes(x = coords.x, y = coords.y, fill = value)) + 
  geom_tile() + 
  facet_wrap(~variable)
```

## Main simulation function

The main function for generating a spatiotemporal system, `generate_spacetime()`, 
incorporates all the functionality from the functions demonstrated above. The
next section will look at some examples of systems that can be created.◘

# 1. Stable (undisturbed) systems

```{r}
bdf.1 <- break_frame(nodes = c(0, 49), slopes = c(0))
sb.1 <- generate_spacebreak(map.dims = c(4, 4), ntime = 50, break.df = bdf.1,
                            covar.pars = list(range = .5),
                            arima.mod = list(ar = .5),
                            sd = 1)

## each pixel through time, with overall trend
breakR:::plot.spacebreak(sb.1)

# Note that the initial values are much more variable than the others, 
# when sd is very low, this is because the burn.in in sparima_sim (and arima.sim) 
# is not working as expected.
```

## 1b (with a trend)

```{r}
bdf.1b <- break_frame(nodes = c(0, 49), slopes = c(.05))
sb.1b <- generate_spacebreak(map.dims = c(4, 4), ntime = 50, break.df = bdf.1b,
                            covar.pars = list(range = .5),
                            arima.mod = list(ar = .5),
                            sd = 1)

## each pixel through time, with overall trend
breakR:::plot.spacebreak(sb.1b)
```


# 2. Seasonal changes

```{r}
bdf.2 <- break_frame(nodes = c(0, 10), slopes = c(0.2))
sb.2 <- generate_spacebreak(map.dims = c(4, 4), break.df = bdf.2,
                            covar.pars = list(range = .5),
                            arima.mod = list(ar = .5),
                            sd = 1,
                            seasonality = list(amp = 2, per = 1), 
                            times = seq(0, 10, by = 1/12))

## each pixel through time, with overall trend
breakR:::plot.spacebreak(sb.2)
```


# 3. Gradual changes

```{r}
bdf.3 <- break_frame(nodes = c(0, 5, 10), slopes = c(.8, -.2))
sb.3 <- generate_spacebreak(map.dims = c(4, 4), break.df = bdf.3,
                            covar.pars = list(range = .5),
                            arima.mod = list(ar = .5),
                            sd = 1, 
                            seasonality = list(amp = 1.2, per = 1), 
                            times = seq(0, 10, by = 1/12))

## each pixel through time, with overall trend
breakR:::plot.spacebreak(sb.3)
```


# 4. Abrupt changes

```{r}
bdf.4 <- break_frame(nodes = c(0, 6 - (1/12), 6, 10), slopes = c(.8, -30, .5))
sb.4 <- generate_spacebreak(map.dims = c(4, 4), break.df = bdf.4,
                            covar.pars = list(range = .5),
                            arima.mod = list(ar = .5),
                            sd = 1, 
                            seasonality = list(amp = 1.2, per = 1), 
                            times = seq(0, 10, by = 1/12))

## each pixel through time, with overall trend
breakR:::plot.spacebreak(sb.4)
```

# Simulate multiple maps and apply fitting functions:

Here, I'll use the same parameters from the "Abrupt Changes" example above, to
simulate 3 different maps and apply a fit function to each map. 

In this example, we'll create our fit function with a simple linear model and
a mixed effects model. However, users interested in breakpoint analyses will
need to write a function that contains each of the breakpoint methods to be
tested. Here is that example fit function:

```{r}
example_fitfun_map <- function(Y, ...){
  # convert Y (matrix) into a long-form dataframe
  data = melt_spacebreak(Y)

  ## fit the models of interest
  model.lm <- stats::lm(value ~ time, data = data) # linear
  model.lmer <- lme4::lmer(value ~ (1|time), data = data) #random effects
  ## any other model that you want goes here ...

  # return the model outputs, even if they differ.
  output <- list(lm = model.lm, lmer = model.lmer)
}
```

Now we'll simulate the 3 maps:

```{r}
sims <- simulate_spacebreak(nmaps = 3, map.dims = c(4, 4), break.df = bdf.4,
                            covar.pars = list(range = .5),
                            arima.mod = list(ar = .5),
                            sd = 1, 
                            seasonality = list(amp = 1.2, per = 1), 
                            times = seq(0, 10, by = 1/12), 
                            fit_FUN = example_fitfun_map)
```

And show a simple output of the results by plotting each of the 3 maps, including
the true trends in red and the fitted model from the simple regression case:

```{r}
for(i in seq_len(length(sims$data))){
  matplot(y = t(sims$data[[i]]), x = sims$time, type = "l", ylab = "Y", xlab = "Time", col = 'grey50',
          main = paste("sim", i))
  lines(sims$f.t ~ sims$time, col = "red", lwd = 2) # true slopes
  
  ## add in linear slope from model
  lines(predict(sims$fits[[1]]$lm, newdata = data.frame(time = sims$time)) ~ sims$time, col = "blue", lty = 2, lwd = 2)
}
```

The linear model does a poor job of approximating the true trends - as expected.
